Towards a standardized evaluation of imputation methodology

### 2022-03-29 Formatted/Formatted.tex
remove some duplicate info from missingness paragraphs in problems and solutions sections

After deciding upon a data generating mechanism, another influential--but often overlooked--decision awaits. Simulation studies on missing data methodolology have the unique option to exclude sampling variance from their evaluations, and only use the missing data generation procedure as source of Monte Carlo variation. After all, we are interested in the missingness and are not considering the noise induced by the sampling mechanism for evaluation in such studies. Therefore, it is not necessary to take sampling variation into account in order to obtain information about a method's ability to handle the missing data problem, or to objectively compare methods on their ability to correct for missingness \citep[see for a detailed discussion][]{vink14}. A lot of published work does not consider this option, while it could have sharpened inconclusive or obfuscated results. The only required change to simulation set-ups is to remove sampling variance from performance evaluation. We highly value the additional flexibility that such a simulation set-up offers. Especially in the case of data transformations it can be challenging to derive the true parametric references to evaluate against. Simply using a single generated complete set in which simulated missingness is induced and evaluating against that generated data set avoids a plethora of procedural problems and computational challenges.

@@

\begin{table}[htb]
\begin{center}
\caption{Missingness mechanisms. [Add PMAR]}
\label{table:mech}
\begin{tabular}{lll}
\hline
Mech. & Interpretation and consequences of the mechanism \\
\hline  
MCAR  & The probability to be missing is the same for all cases. In other words, the missing \\
      & values are missing at random and the observed values are observed at random. \\
MAR   & The probability to be missing is the same within groups of cases defined by the \\
      & observed data only. The essence is that observed relations in the data inform the \\
      & missingness, such that MCAR may be assumed within the observed groups. \\
MNAR  & The probability to be missing depends on unobserved aspects of the data. The \\
      & cause of the missingness is unknown and cannot be inferred from the observed data. \\
      & This is considered non-ignorable missingness \citep[see e.g.][]{rubi76}. \\
\hline
\end{tabular}
\end{center}
\end{table}

The missingness mechanism is usually assumed to be random (MAR) or completely random (MCAR). The validity of the assumptions becomes increasingly important when the missingness proportion increases.

@@

All imputation models should be congenial, but this is difficult to test and often ignored. Checks of imputation model fit are often omitted. Ignoring congeniality may yield sub-optimal imputations, which may disadvantage certain methods compared to others. Same goes for convergence: FCS is iterative and requires algorithmic convergence, but this is typically evaluated through visual inspection which is unfeasible in simulation studies. Problems with convergence may be even more severe: some methods do not produce imputations at all for certain conditions, while at other times the imputation algorithm is cut off too soon, and performance will be under-estimated.]

@@

The scope of the study also encompasses the general simulation design. Ideally, one should perform fully factorial design, where each simulation condition is varied against all other conditions. This is important because there are known interaction effects between e.g. missingness proportions and missingness mechanisms. If multiple imputation methods are under evaluation (and/or parallel computing is used), it may be wise to split-up the simulation workflow into separate parts, such that each imputation method is applied to the same incomplete dataset within every simulation repetition.

@@

The simulator should choose their data generating mechanism(s) and estimand(s) in line with the study scope. Model-based data-generating mechanisms have the advantage of flexibility and precision, since data are generated from a probability distribution or statistical model and the true theoretical parameters can be known. A design-based approach is often used in situations where a probability distribution is not available, or where real-life data structures are of interest. A benefit of design-based simulation is the ability to use real-life observed data structures.

@@

[Missingness mechanisms:] 
First, one should always consider MCAR missingness, i.e. the scenario where the missing values are missing at random and the observed values are observed at random. % Under MCAR, the statistical properties of the observed data given the missing data are known and any imputation routine that cannot at least mimic the performance of the observed data inference, should be deemed inefficient in the scope of the simulation. 
Next, missing data should be induced conform a model that is dependent on the observed data. A straightforward technique for inducing different forms of univariate MAR missingness is described in \citet[][p. 63]{fimd} and a generalization to multivariate MAR missingness can be found in \citet{ampute}. If the missingness is to be induced in longitudinal data, autoregressive MAR models \citep[e.g. cf.][model 2 and model 3]{shara2015randomly} can be useful.
[Although often ignored, it can be argued that MNAR is the more likely mechanism for real-life missingness scenarios. It would be wise to include the mechanism in your simulation study, or perform sensitivity analyses.] An indication of the validity of the obtained inference, given that the assumed missingness mechanism is suspected to be invalid, may be obtained by performing sensitivity analysis \citep[see e.g.][part 5]{molenberghs2014handbook}.

[Missingness type:] 
Third, it is advisable to investigate varying shapes of MAR missingness to achieve a more realistic indication of the robustness of the imputation performance across the range of random missingness. The eefects of different forms of MAR mechanisms are described in \citet{scho18}.% Given the simulated data distributions, one random missingness model may be far more disastrous to the observed information than another model. This may influence the performance of some (but not necessarily all) imputation routines. For example, inference from hot-deck techniques such as predictive mean matching \citep{little1988missing, rubin1986statistical} may be more severely impacted by large amounts of one-tailed missingness than inference from parametric techniques.
It would be a shame to overlook results due to the focus on a single MAR mechanism.

[Missingness proportion: ]
Fourth, the amount of missingness must be varied. Remember that missingness is only ignorable under MAR when the parameter of the data is distinct and a-priori independent from the parameter of the missing data process. Under MAR missingness we assume that we may use the observed data to make inferences about the joint (observed and unobserved) data. The dependency of the procedure on the assumption under which we obtain inference is only influenced by the amount of missingness. If there is no missingness--or if there is no data, for that matter--the inference does not depend on the assumption. Alternatively, the validity of assumptions become increasingly important when the missingness increases. Since we control the MAR mechanism, the assumption under which we may solve the missing data problem should hold and it is only fair to assess performance under stringent missingness conditions. We, therefore, propose to evaluate imputation methodology under 10\%, 25\% and 50\% univariate missingness see Table 2. Although we limit the focus here to ignorable non-response, the suggested proportions are equally applicable to simulations under non-ignorable non-response.

\begin{table}[htb]
\begin{center}
\caption{Suggested univariate missingness proportions.}
\begin{tabular}{lll}
\hline
Prop. & Reasoning for suggested missingness proportion \\
\hline  
10\%  & Depending on the size of the data, this percentage can be considered as a lower \\
      & bound of realistic evaluation. Anything less than 10\% may be of little influence on \\
      & the true data inference. Performance of a missing data method should at least be \\
      & acceptable for most missing data problems. \\ % & \\
25\%  & This is a fair amount of missingness and will, depending on the observed data \\
      & information, have a noticeable influence on the completed data inference. When \\
      & compared to the condition with 10\% missingness, the inference obtained under \\
      & 25\% missingness should be less certain (i.e. confidence/credibility interval width \\
      & should increase), but estimates should still be properly covered and the statistical \\
      & properties of the missing data method should be sound. In practice, at least to our \\
      & at least to my experience in social sciences and official statistics, 25\% \\
      & univariate missingness can easily be considered as a realistic missingness percentage. \\ % & \\
50\%  & Performance under 50\% percent simulated missingness will most likely be impacted \\ 
      & severely. Depending on how the missingness mechanism interacts with the simulated \\
      & data, some imputation techniques may yield estimates that are under-covered such that \\
      & the completed data inference should not be deemed valid anymore. If a method yields \\
      & acceptable inference under 50\% MAR missingness, we can determine that the statistical \\
      & properties of the imputation methodology are sound. \\
\hline
\end{tabular}
\end{center}
\end{table}

>>>>

After deciding upon a data generating mechanism, another influential--but often overlooked--decision awaits. Simulation studies on missing data methodolology have the unique option to exclude sampling variance from their evaluations, and only use the missing data generation procedure as source of Monte Carlo variation. After all, we are interested in the missingness and are not considering the noise induced by the sampling mechanism for evaluation in such studies. Therefore, it is not necessary to take sampling variation into account in order to obtain information about a method's ability to handle the missing data problem, or to objectively compare methods on their ability to correct for missingness \citep[see for a detailed discussion][]{vink14}. The only required change to simulation set-ups is to remove sampling variance from performance evaluation. A lot of published work does not consider this option, while it could have sharpened inconclusive or obfuscated results.

@@

The missingness mechanism is usually assumed to be random (MAR) or completely random (MCAR).

\begin{table}[htb]
\begin{center}
\caption{Missingness mechanisms. [Add PMAR]}
\label{table:mech}
\begin{tabular}{lll}
\hline
Mech. & Interpretation and consequences of the mechanism \\
\hline  
MCAR  & Missing completely at random: The probability to be missing is the same for all \\
      & cases. In other words, the missing values are missing at random and the observed \\
      & values are observed at random. \\
MAR   & Missing at random: The probability to be missing is the same within groups of \\
      & cases defined by the observed data only. The essence is that observed relations in \\
      & the data inform the missingness, such that MCAR may be assumed within the \\ 
      & observed groups. \\
MNAR  & Missing not at random: The probability to be missing depends on unobserved \\
      & aspects of the data. The cause of the missingness is unknown and cannot be \\ 
      & inferred from the observed data. This is considered non-ignorable missingness. \\
\hline
\end{tabular}
\end{center}
\end{table}

@@

[All imputation models should be congenial, but this is difficult to test and often ignored. Checks of imputation model fit are often omitted. Ignoring congeniality may yield sub-optimal imputations, which may disadvantage certain methods compared to others. Same goes for convergence: FCS is iterative and requires algorithmic convergence, but this is typically evaluated through visual inspection which is unfeasible in simulation studies. Problems with convergence may be even more severe: some methods do not produce imputations at all for certain conditions, while at other times the imputation algorithm is cut off too soon, and performance will be under-estimated.]

@@

The scope of the study also encompasses the general simulation design. Ideally, one should perform fully factorial design, where each simulation condition is varied against all other conditions. This may be important because of known interaction effects, e.g. the validity of the missingness mechanism assumptions becomes increasingly important when the missingness proportion increases. If multiple imputation methods are under evaluation (and/or parallel computing is used), it may be wise to split-up the simulation workflow into separate parts, such that each imputation method is applied to the same incomplete dataset within every simulation repetition.

@@

The simulator should choose their data generating mechanism(s) and estimand(s) in line with the study scope. Model-based data-generating mechanisms have the advantages of flexibility and precision since data are generated from a known statistical model so the true theoretical parameters can be derived. A design-based approach is often used in situations where a probability distribution is not available, or where real-life data structures are of interest. A benefit of design-based simulation is the ability to use real-life observed data structures.

@@

Missingness should be induced to generate different missing data conditions. [Discussed are missingness mechanisms incl different types (see Table \ref{table:mech} for definitions), missingness patterns incl different missingness proportions (see Table \ref{table:prop} for suggested proportions).]

First, one should always consider MCAR missingness [as reference, because all imputation methods should yield valid inference under MCAR and we can use this mechanism to check imputation model fit] %, i.e. the scenario where the missing values are missing at random and the observed values are observed at random. % Under MCAR, the statistical properties of the observed data given the missing data are known and any imputation routine that cannot at least mimic the performance of the observed data inference, should be deemed inefficient in the scope of the simulation. 
Next, missing data should be induced conform a model that is dependent on the observed data. A straightforward technique for inducing different forms of univariate MAR missingness is described in \citet[][p. 63]{fimd} and a generalization to multivariate MAR missingness can be found in \citet{ampute}. If the missingness is to be induced in longitudinal data, autoregressive MAR models \citep[e.g. cf.][model 2 and model 3]{shara2015randomly} can be useful.
Third, it is advisable to investigate varying shapes of MAR missingness to achieve a more realistic indication of the robustness of the imputation performance across the range of random missingness. The effects of different forms of MAR mechanisms are described in \citet{scho18}.% Given the simulated data distributions, one random missingness model may be far more disastrous to the observed information than another model. This may influence the performance of some (but not necessarily all) imputation routines. For example, inference from hot-deck techniques such as predictive mean matching \citep{little1988missing, rubin1986statistical} may be more severely impacted by large amounts of one-tailed missingness than inference from parametric techniques. It would be a shame to overlook results due to the focus on a single MAR mechanism. 
[Although often ignored, it can be argued that MNAR is the more likely mechanism for real-life missingness scenarios. It would be wise to include the mechanism in your simulation study, or perform sensitivity analyses.] An indication of the validity of the obtained inference, given that the assumed missingness mechanism is suspected to be invalid, may be obtained by performing sensitivity analysis \citep[see e.g.][part 5]{molenberghs2014handbook}.

Fourth, the amount of missingness must be varied. Remember that missingness is only ignorable under MAR when the parameter of the data is distinct and a-priori independent from the parameter of the missing data process. Under MAR missingness we assume that we may use the observed data to make inferences about the joint (observed and unobserved) data. The dependency of the procedure on the assumption under which we obtain inference is only influenced by the amount of missingness. If there is no missingness--or if there is no data, for that matter--the inference does not depend on the assumption. Alternatively, the validity of assumptions become increasingly important when the missingness increases. Since we control the MAR mechanism, the assumption under which we may solve the missing data problem should hold and it is only fair to assess performance under stringent missingness conditions. We, therefore, propose to evaluate imputation methodology under 10\%, 25\% and 50\% univariate missingness see Table 2. Although we limit the focus here to ignorable non-response, the suggested proportions are equally applicable to simulations under non-ignorable non-response.

\begin{table}[htb]
\begin{center}
\caption{Suggested univariate missingness proportions.}
\label{table:prop}
\begin{tabular}{lll}
\hline
Prop. & Reasoning for suggested missingness proportion \\
\hline  
10\%  & Depending on the size of the data, this percentage can be considered as a lower \\
      & bound of realistic evaluation. Anything less than 10\% may be of little influence on \\
      & the true data inference. Performance of a missing data method should at least be \\
      & acceptable for most missing data problems. \\ % & \\
25\%  & This is a fair amount of missingness and will, depending on the observed data \\
      & information, have a noticeable influence on the completed data inference. When \\
      & compared to the condition with 10\% missingness, the inference obtained under \\
      & 25\% missingness should be less certain (i.e. confidence/credibility interval width \\
      & should increase), but estimates should still be properly covered and the statistical \\
      & properties of the missing data method should be sound. In practice, at least to our \\
      & at least to my experience in social sciences and official statistics, 25\% \\
      & univariate missingness can easily be considered as a realistic missingness percentage. \\ % & \\
50\%  & Performance under 50\% percent simulated missingness will most likely be impacted \\ 
      & severely. Depending on how the missingness mechanism interacts with the simulated \\
      & data, some imputation techniques may yield estimates that are under-covered such that \\
      & the completed data inference should not be deemed valid anymore. If a method yields \\
      & acceptable inference under 50\% MAR missingness, we can determine that the statistical \\
      & properties of the imputation methodology are sound. \\
\hline
\end{tabular}
\end{center}
\end{table}
